P3064R1: How to Avoid OOTA Without Really Trying

### 2024-05-13 ootavacuous.tex
Fixes to commits a186bdfbf6fe and 9a8e75ed86a7
Commits a186bdfbf6fe ("Section 4.3: Move forward references to a
footnote") and 9a8e75ed86a7 ("May 12 feedback from Alan Stern") could
stand a little improvement.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

A noteworthy aspect of this transformation is that it invents an atomic load:
The original form of the code reads \co{x} only once,
whereas the transformed code reads it twice.
Therefore we can rule out transformations like this one by insisting
the compiler not invent (or duplicate) atomic loads.\footnote{
	This issue is discussed more fully in
	Appendix~\ref{app:Inventing Atomic Loads},
	which also discusses other examples in which invented atomic
	loads lead to incorrect behavior, that is, behavior that is not
	exhibited by the abstract machine.
	See also the ``Invented Atomic Loads'' paragraph of
        Section~\ref{sec:Constraints of the Standard}.}
In fact, we will require that atomic accesses be treated as
``quasi volatile'', in that the compiler is allowed to omit,
merge, or reorder them but not invent them.

@@

\paragraph{Invented Atomic Loads}
\begin{listing}[tbp]
\scriptsize
\begin{verbatim}
 1 struct foo {
 2   int a;
 3   int b;
 4 };
 5 _Atomic struct foo *globalfoop;
 6
 7 void bar()
 8 {
 9   int a;
10   int b;
11   struct foo *fp;
12
13   fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
14   a = fp->a;
15   // fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
16   b = fp->b;
17   do_something_with(a, b);
18 }
\end{verbatim}
\caption{Example Duplicated Load}
\label{lst:Example Duplicated Load}
\end{listing}
Although one can argue that it is functionally correct to invent atomic
loads as long as the loaded value is not used, doing so can result in
extra cache misses for no good purpose.
And duplicating an atomic load can also lead to incorrect results if done
carelessly.
To see this, consider
Listing~\ref{lst:Example Duplicated Load}.
Suppose that the load on line~13 were duplicated, for example, as if the
compiler had uncommented line~15.
uncommenting line~15.
Then the values of \co{a} and \co{b} passed to line~17's call to
\co{do_something_with()} might be from different instances of the
\co{foo} structure, something that \co{do_something_with()} probably
isn't prepared to deal with and that represents incorrect behavior,
that is, behavior that the abstract machine does not exhibit.
See
Appendix~\ref{app:Inventing Atomic Loads}
for more details on this example and other examples of invented
atomic loads.

@@

But if the compiler is permitted to invent atomic loads, then compiled
code might not respect this guarantee, thus (incorrectly) deviating
from the compiled code.
This issue is evident in this transformed version of the above code:
\begin{quote}
\begin{verbatim}
int r0 = x;
int invented = x;
int r1 = r0 * r0 + 2 * invented + 1;
\end{verbatim}
\end{quote}
For example, if the load into \co{r0} happened when the value of \co{x}
was zero and the load into \co{invented} happened when the value of
\co{x} was 10, then the value of \co{r1} will be 21, which is not
a perfect square.

@@

Similar observations apply to atomic loads from pointers.
For example, consider
Listing~\ref{lst:Example Duplicated Load} on
page~\pageref{lst:Example Duplicated Load},
in which the abstract machine guarantees that line~17 passes two
fields from the same structure to \co{do_something_with()}.
This guarantee will not always be met if the compiler is permitted to
invent atomic loads such as that contained in the commented-out line~15.
\begin{listing}[tbp]
\scriptsize
\begin{verbatim}
 1 struct foo {
 2   int a;
 3   int b;
 4 };
 5 _Atomic struct foo *globalfoop;
 6
 7 void bar()
 8 {
 9   int a;
10   int b;
11   struct foo *fp;
12
13   fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
14   do_something(fp);
15   fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
16   a = fp->a;
17   // fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
18   b = fp->b;
19   do_something_with(a, b);
20 }
\end{verbatim}
\caption{Example Merging of Duplicated Loads}
\label{lst:Example Merging of Duplicated Loads}
\end{listing}
But suppose that there was an earlier load from \co{globalfoop}, as
shown on lines~13 and~14 in
Listing~\ref{lst:Example Merging of Duplicated Loads}.
Then, assuming that \co{do_something()} is synchronization-free, the
second load on line~15 if that listing could be removed without resulting
in incorrect behavior.

>>>>

A noteworthy aspect of this transformation is that it invents an atomic load:
The original form of the code reads \co{x} only once,
whereas the transformed code reads it twice.
Therefore we can rule out transformations like this one by insisting
the compiler not invent (or duplicate) atomic loads.\footnote{
	This issue is discussed more fully in
	Appendix~\ref{app:Inventing Atomic Loads},
	which includes other examples of invented atomic
	loads leading to incorrect behavior, that is, behavior not
	exhibited by the abstract machine.
	See also the ``Invented Atomic Loads'' paragraph of
        Section~\ref{sec:Constraints of the Standard}.}
In fact, we will require that atomic accesses be treated as
``quasi volatile'', in that the compiler is allowed to omit,
merge, or reorder them but not invent them.

@@

\paragraph{Invented Atomic Loads}
\begin{listing}[tbp]
\scriptsize
\begin{verbatim}
 1 struct foo {
 2   int a;
 3   int b;
 4 };
 5 _Atomic struct foo *globalfoop;
 6
 7 void bar()
 8 {
 9   int a;
10   int b;
11   struct foo *fp;
12
13   fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
14   a = fp->a;
15   // fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
16   b = fp->b;
17   do_something_with(a, b);
18 }
\end{verbatim}
\caption{Example Duplicated Load}
\label{lst:Example Duplicated Load}
\end{listing}
Although one can argue that it is functionally correct to invent atomic
loads as long as the loaded value is not used, doing so can result in
extra cache misses for no good purpose.
And duplicating an atomic load can also lead to incorrect results if done
carelessly.
To see this, consider
Listing~\ref{lst:Example Duplicated Load}.
Suppose that the load on line~13 were duplicated, for example, as if the
compiler had uncommented line~15.
Then the values of \co{a} and \co{b} passed to line~17's call to
\co{do_something_with()} might be from different instances of the
\co{foo} structure, something that \co{do_something_with()} probably
isn't prepared to deal with and that represents incorrect behavior,
that is, behavior that the abstract machine does not exhibit.
See
Appendix~\ref{app:Inventing Atomic Loads}
for more details on this example and other examples of invented
atomic loads.

@@

But if the compiler is permitted to invent atomic loads then the compiled
code might not respect this guarantee, thus incorrectly deviating
from the abstract machine.
This issue is evident in a transformed version of the above code:
\begin{quote}
\begin{verbatim}
int r0 = x;
int invented = x;
int r1 = r0 * r0 + 2 * invented + 1;
\end{verbatim}
\end{quote}
For example, if the load into \co{r0} happened when the value of \co{x}
was zero and the load into \co{invented} happened when the value of
\co{x} was 10, then the value of \co{r1} will be 21, which is not
a perfect square.

@@

Similar observations apply to atomic loads from pointers.
For example, consider
Listing~\ref{lst:Example Duplicated Load} on
page~\pageref{lst:Example Duplicated Load},
in which the abstract machine guarantees that line~17 passes two
fields from the same structure to \co{do_something_with()}.
This guarantee will not always be met if the compiler is permitted to
invent atomic loads such as that contained in the commented-out line~15.
\begin{listing}[tbp]
\scriptsize
\begin{verbatim}
 1 struct foo {
 2   int a;
 3   int b;
 4 };
 5 _Atomic struct foo *globalfoop;
 6
 7 void bar()
 8 {
 9   int a;
10   int b;
11   struct foo *fp;
12
13   fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
14   do_something(fp);
15   fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
16   a = fp->a;
17   // fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
18   b = fp->b;
19   do_something_with(a, b);
20 }
\end{verbatim}
\caption{Example Merging of Duplicated Loads}
\label{lst:Example Merging of Duplicated Loads}
\end{listing}
But suppose that there was an earlier load from \co{globalfoop}, as
shown on lines~13 and~14 in
Listing~\ref{lst:Example Merging of Duplicated Loads}.
Then, assuming that \co{do_something()} is synchronization-free, the
second load on line~15 of that listing could be removed without resulting
in incorrect behavior.
